---
title: "TCGAbioinks: examples using GDC api"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
knitr::opts_knit$set(progress = FALSE)
```


TCGAbiolinks is able to access The National Cancer Institute (NCI) Genomic Data Commons (GDC) thourogh their  
[GDC Application Programming Interface (API)](https://gdc.cancer.gov/developers/gdc-application-programming-interface-api) 
to search, download and prepare relevant data for data analysis in R. 

You may install the stable version from Biocondcutor, or the development version using devtools::install_github('BioinformaticsFMRP/TCGAbiolinks').

Please use [Github issues](https://github.com/BioinformaticsFMRP/TCGAbiolinks/issues) if you want to file bug reports or feature requests.

To run the examples please load the following libraries:
```{r sourceCode, message=FALSE, warning=FALSE}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)
library(DT)
```

# Session info
```{r}
version
packageVersion("TCGAbiolinks")
```
