---
title: "Introduction"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(progress = FALSE)
```


TCGAbiolinks is able to access The National Cancer Institute (NCI) Genomic Data Commons (GDC) thorough its  
[GDC Application Programming Interface (API)](https://gdc.cancer.gov/developers/gdc-application-programming-interface-api) 
to search, download and prepare relevant data for analysis in R. 

# Installation
You can install the stable version from [Bioconductor](http://bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html), or the development version using:
```{r message=FALSE, warning=FALSE, eval=FALSE}
devtools::install_github('BioinformaticsFMRP/TCGAbiolinks')
```

# Question and issues
Please use [Github issues](https://github.com/BioinformaticsFMRP/TCGAbiolinks/issues) if you want to file bug reports or feature requests.

# Required libraries

The examples in this tutorial use the following libraries: 
```{r message=FALSE, warning=FALSE, include=TRUE}
library(TCGAbiolinks)
library(dplyr)
library(DT)
```

# Session info
```{r}
version
packageVersion("TCGAbiolinks")
```
