---
title: "TCGAbiolinks: Searching and downloading mutation files"
date: "`r BiocStyle::doc_date()`"
vignette: >
  %\VignetteIndexEntry{"TCGAbiolinks: Mutation data"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(progress = FALSE)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)
library(DT)
```


**TCGAbiolinks** has provided a few functions to download mutation data from GDC.
There are two options to download the data:

1. Use `GDCquery_Maf` which will download MAF aligned against hg38
2. Use `GDCquery`, `GDCdownload` and `GDCpreprare` to downoad MAF aligned against hg19


---

# Mutation data (hg38)

This exmaple will download MAF (mutation annotation files) for variant calling pipeline muse.
Pipelines options are: muse, varscan2, somaticsniper, mutect. For more information please access
[GDC docs](https://docs.gdc.cancer.gov/Data/Bioinformatics_Pipelines/DNA_Seq_Variant_Calling_Pipeline/).

```{r results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval = FALSE}
acc.maf <- GDCquery_Maf("ACC", pipelines = "muse")
```
```{r  echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE}
# Only first 50 to make render faster
datatable(acc.maf[1:50,],
              filter = 'top',
              options = list(scrollX = TRUE, keys = TRUE, pageLength = 5), 
              rownames = FALSE)
```

# Mutation data (hg19)

This exmaple will download MAF (mutation annotation files) aligned against hg19 (Old TCGA maf files)


```{r results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval = FALSE}
query.maf.hg19 <- GDCquery(project = "TCGA-CHOL", 
                           data.category = "Simple nucleotide variation", 
                           data.type = "Simple somatic mutation",
                           access = "open", 
                           legacy = TRUE)
```
```{r  echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE}
# Check maf availables
datatable(select(getResults(query.maf.hg19),-contains("cases")),
          filter = 'top',
          options = list(scrollX = TRUE, keys = TRUE, pageLength = 10), 
          rownames = FALSE)
```
```{r results = 'hide', echo=TRUE, message=FALSE, warning=FALSE, eval = FALSE}
query.maf.hg19 <- GDCquery(project = "TCGA-CHOL", 
                           data.category = "Simple nucleotide variation", 
                           data.type = "Simple somatic mutation",
                           access = "open", 
                           file.type = "bcgsc.ca_CHOL.IlluminaHiSeq_DNASeq.1.somatic.maf",
                           legacy = TRUE)
GDCdownload(query.maf.hg19)
maf <- GDCprepare(query.maf.hg19)
```
```{r  echo = TRUE, message = FALSE, warning = FALSE, eval = F}
# Only first 50 to make render faster
datatable(maf[1:50,],
          filter = 'top',
          options = list(scrollX = TRUE, keys = TRUE, pageLength = 5), 
          rownames = FALSE)
```

