---
title: "Working with TCGAbiolinks package"
author: " Antonio Colaprico, Catharina Olsen, Luciano Garofano, Tiago Chedraoui Silva, Davide Garolini, Claudia Cava, Isabella Castiglioni, Houtan Noushmehr, Gianluca Bontempi, Michele Ceccarelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This document describes how to search/download analyze tcga data using TCGAbiolinks package.

## TCGA query
 You can easily search tcga samples using the `tcgaQuery` function.
 Using a summary of filters as in TCGA portal, the function works with the following 
 parameters:

* tumor
* platform
* added.since
* added.up.to
* listSample
* level
* center

Some examples of search are shown below
```{r, eval = FALSE}

library(TCGAbiolinks)
query <- TCGAQuery(tumor = "brca",level = 3)

querySamples <- samplesfilter(query)

query <- TCGAQuery(tumor = "brca",level = 3, platform = "IlluminaHiSeq_RNASeqV2")

```

Obs: As these terms needs to be exact we validate user input before searching.

## TCGA download
You can easily download roadmap data using the `tcga.download` function.
 
```{r, eval = FALSE}
# get all samples from the query and save into tcga folder
# samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results
# samples to normalize later
TCGADownload(query,path = "data",type ="rsem.genes.results")
```

Obs: The function will structure the folders to save the data as:
_Path given by the user/Experiment folder_
