---
title: "Working with TCGA inside biOmics"
author: "Tiago Chedraoui Silva"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This document describes how to search/download tcga data using biOmics package.

## TCGA search
 You can easily search tcga samples using the `tcgaQuery` function.
 Using a summary of roadmap datatabase, the function works with the following 
 parameters:

* tumor
* platform
* added.since
* added.up.to
* listSample
* level
* center

Some examples of search are shown below
```{r, eval = FALSE}
query <- tcgaQuery(tumor = "gbm",
                    added.since = "01/01/2013",
                    added.up.to = "06/01/2013")

query <- tcgaQuery(tumor = c("gbm","lgg"),
                    platform = c("HumanMethylation450","HumanMethylation27"))

query <- tcgaQuery(tumor = "gbm", platform = "bio")

query <- tcgaQuery(tumor = "gbm", platform = "HumanMethylation450",
                    level = "3")

query <- tcgaQuery(samples = "TCGA-61-1743-01A-01D-*")

query <- tcgaQuery(samples = "TCGA-61-1743-01A-01D-0649-04")

query <- tcgaQuery(samples = "TCGA-61-1743-01A-01D-0649-04", level = 3)

query <- tcgaQuery(samples = "TCGA-61-1743-01A-01D-0649-04",
                    tumor = "OV", platform = "CGH-1x1M_G4447A")
```

Obs: As these terms needs to be exact we validate user input before searching.

## Roadmap download
You can easily download roadmap data using the `tcga.download` function.
 
```{r, eval = FALSE}
# get all sampples from the query and save into tcga folder
tcgaDownload(query,path="tcgaData")
```

Obs: The function will structure the folders to save the data as:
_Path given by the user/Experiment folder_
