---
title: "Working with TCGAbiolinks package"
author: " Antonio Colaprico, Tiago Chedraoui Silva, Luciano Garofano, Catharina Olsen, Davide Garolini, Claudia Cava, Isabella Castiglioni, Houtan Noushmehr, Gianluca Bontempi, Michele Ceccarelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This document describes how to search, download and analyze TCGA data using the TCGAbiolinks package.

## TCGA query
 You can easily search TCGA samples using the `tcgaQuery` function.
 Using a summary of filters as used in the TCGA portal, the function works with the following 
 parameters:

* tumor
* platform
* added.since
* added.up.to
* listSample
* level
* center

Some search examples are shown below
```{r, eval = FALSE}

library(TCGAbiolinks)
query <- TCGAQuery(tumor = "brca",level = 3)
querySamples <- samplesfilter(query)
query <- TCGAQuery(tumor = "brca",level = 3, platform = "IlluminaHiSeq_RNASeqV2")

```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)

query <- TCGAQuery(tumor = "brca",level = 3)
querySamples <- samplesfilter(query)

matSamples <- matrix(0,length(querySamples),length(querySamples))
colnames(matSamples) <- names(querySamples)
rownames(matSamples) <- names(querySamples)

for ( i in 1: nrow(matSamples)){
    for( j in 1: nrow(matSamples)){
             matSamples[i,j] <- length(intersect(querySamples[[i]], querySamples[[j]]))
    }
    }


colnames(matSamples) <- paste("p",1:length(names(querySamples)), names(querySamples),sep=".")
rownames(matSamples) <- paste("p",1:length(names(querySamples)), names(querySamples),sep=".")
colnames(matSamples) <- substr( colnames(matSamples), 1,4)

knitr::kable(matSamples, digits = 2, caption = "Table common samples among platforms from TCGAQuery", row.names = TRUE)
```

## TCGA query, download, prepare list of a barcode
You can easily search TCGA samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}

# You can define a list of samples to query and download providing relative TCGA barcodes.
listSamples <- c("TCGA-33-6737-01","TCGA-33-6737-11","TCGA-43-6143-01","TCGA-43-6143-11",
                 "TCGA-43-6647-01","TCGA-43-6647-11","TCGA-43-6771-01","TCGA-43-6771-11")

# Query all available platforms with a list of barcode 
query <- TCGAQuery(samples = listSamples)

# Query platform IlluminaHiSeq_RNASeqV2 with a list of barcode 
query <- TCGAQuery(tumor = "lusc", samples = listSamples, 
  platform = "IlluminaHiSeq_RNASeqV2", level = "3")

# Download a list of barcodes with platform IlluminaHiSeq_RNASeqV2
TCGADownload(query, path = "dataLusc", type = "rsem.genes.results", samples = listSamples)

# Prepare expression matrix with gene id in rows and samples (barcode) in columns
# rsem.genes.results as values
LuscMatrix <- TCGAPrepare(query,"dataLusc")
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
dataGE <- dataBRCA[sample(rownames(dataBRCA),5),sample(colnames(dataBRCA),3)]

knitr::kable(dataGE, digits = 2, caption = "Matrix Gene expression (genes in rows and samples in columns)",row.names = FALSE)
```


## TCGA download
You can easily download data using the `tcga.download` function.
 
```{r, eval = FALSE}
# get all samples from the query and save them in the tcga folder
# samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results
# samples to normalize later
TCGADownload(query, path = "data", type = "rsem.genes.results")

```

Comment: The function will structure the folders to save the data as:
_Path given by the user/Experiment folder_

## TCGA download clinic data
 
```{r, eval = FALSE}

# Function clinic
# clinic(cancer,clinical_data_type)
# input: cancer, clinical_data_type 
# cancer = a character vector indicating cancer type to download
# Options include:
# ACC,BLCA,BRCA,CESC,COAD,DLBC,ESCA,GBM
# HNSC,KICH,KIRC,KIRP,LAML,LGG,LIHC,LUAD,
# LUSC,OV,PAAD,PRAD,READ,SARC,SKCM,STAD,THCA
# UCEC,UCS

# For information about cancer types: https://tcga-data.nci.nih.gov/tcga/
#
# clinical_data_type = a character vector indicating the types of 
# clinical data 
# 
# Options include:
# biospecimen_aliquot, biospecimen_analyte, biospecimen_cqcf, biospecimen_diagnostic_slides,
# biospecimen_normal_control, biospecimen_portion, biospecimen_protocol,
# biospecimen_sample, biospecimen_shipment_portion, biospecimen_slide,
# biospecimen_tumor_sample, clinical_cqcf, clinical_drug, clinical_follow_up_v1.5,
# clinical_follow_up_v2.1, clinical_follow_up_v4.0, clinical_follow_up_v4.0_nte,
# clinical_nte, clinical_omf_v4.0, clinical_patient, clinical_radiation 
# 
                       
clinic("gbm","clinical_patient")

```


## TCGA Downstream Analysis TCGAprepare, TCGAanalyze, TCGAvisualize
You can prepare gene expression from TCGA data using the `tcga.prepare` function.
 
```{r, eval = FALSE}
# Downstream analysis using gene expression data  
# TCGA samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results

# save(DAVID_BP_matrix, DAVID_CC_matrix, DAVID_MF_matrix, IPAGenes, listIPA_pathways,
#   file = "dataEnrichmentAnalysis.rda")
# save(dataBRCA, geneInfo , file = "dataGeneExpression.rda")
library(TCGAbiolinks)

# normalization of genes
dataNorm <- TCGAbiolinks::RnaSeqNormalization(dataBRCA, geneInfo)

# quantile filter of genes
dataFilt <- RnaSeqFilt(dataNorm, 0.25)

# selection of normal samples "NT"
samplesNT <- MultiSampleTypes(colnames(dataFilt), typesample = c("NT"))

# selection of tumor samples "TP"
samplesTP <- MultiSampleTypes(colnames(dataFilt), typesample = c("TP"))


```

## TCGA analyze and visualize (Diff.expr.analysis (DEA))
You can analyze gene expression from TCGA data using the `tcga.analyze` function.
 
```{r, eval = FALSE}
# Downstream analysis using gene expression data  
# TCGA samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results
# save(dataBRCA, geneInfo , file = "dataGeneExpression.rda")
library(TCGAbiolinks)

# Diff.expr.analysis (DEA)
dataDEGs <- DEArnaSEQ(dataFilt[,samplesNT], dataFilt[,samplesTP], "Normal", "Tumor")

# DEGs filter by abs(logFC) >=1
dataDEGsFilt <- dataDEGs[abs(dataDEGs$logFC) >= 1,]

# DEGs table with expression values in normal and tumor samples
dataDEGsFiltLevel <- CreateTabLevel(dataDEGsFilt,"Tumor","Normal",dataFilt[,samplesTP],dataFilt[,samplesNT])

```
The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
dataDEGsFiltLevel$FDR <- format(dataDEGsFiltLevel$FDR, scientific = TRUE)
knitr::kable(dataDEGsFiltLevel[1:10,], digits = 2, caption = "Table DEGs after DEA",row.names = FALSE)
```



## TCGA analyze and visualize (Enrichment Analysis)
You can analyze gene expression from TCGA data using the `tcga.analyze` function.
 
```{r, eval = FALSE}

# Enrichment Analysis EA
# Gene Ontology (GO) and Pathway enrichment by DEGs list
Genelist <- rownames(dataDEGsFiltLevel)

TimeUse(ansEA <- EAcomplete(TFname="DEA genes Normal Vs Tumor",Genelist))

# Enrichment Analysis EA (TCGAVisualize)
# Gene Ontology (GO) and Pathway enrichment IPA barPlot

IPAbarplot(tf = rownames(ansEA$ResBP), 
           GOBPTab = ansEA$ResBP,
           GOCCTab = ansEA$ResCC,
           GOMFTab = ansEA$ResMF,
           PathTab = ansEA$ResPat,
           nRGTab = Genelist, 
           nBar = 10)




```
The result is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("EAplot.png")
grid.raster(img)
```

## TCGA analyze (Enrichment Analysis)
You can analyze gene expression from TCGA data using the `tcga.analyze` function.
 
```{r, eval = FALSE}
library(TCGAbiolinks)

# normalization of genes
dataNorm <- TCGAbiolinks::RnaSeqNormalization(dataBRCA, geneInfo)

# quantile filter of genes
dataFilt <- RnaSeqFilt(dataNorm, 0.25)

# Principal Component Analysis plot for ntop selected DEGs
plotPCAforGroups(dataFilt,dataDEGsFiltLevel, ntopgenes = 200)

# boxplot of normalized data
#sampleGenes <- rownames(dataDEGsFilt[dataDEGsFilt$logFC >=1,])[1:20]
#boxplot(log(dataBRCA[sampleGenes,]), las = 2)
#boxplot(log(dataFilt[sampleGenes,]), las = 2)


```
The result is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("PCAtop200DEGs.png")
grid.raster(img)
```



## TCGA investigate
Find most studied TFs in pubmed related to a specific cancer, disease, or tissue
 
```{r, eval = FALSE}
# First perform DEGs with TCGAanalyze
# See previous section 
library(TCGAbiolinks)

# Select only transcription factors (TFs) from DEGs
TFs <- IPAGenes[IPAGenes$Family =="transcription regulator",]
TFs_inDEGs <- intersect(TFs$Gene, dataDEGsFiltLevel$mRNA )
dataDEGsFiltLevelTFs <- dataDEGsFiltLevel[TFs_inDEGs,]

# Order table DEGs TFs according to Delta decrease
dataDEGsFiltLevelTFs <- dataDEGsFiltLevelTFs[order(dataDEGsFiltLevelTFs$Delta,decreasing = TRUE),]

# Find Pubmed of TF studied related to cancer
tabDEGsTFPubmed <- TCGAinvestigate("breast", dataDEGsFiltLevelTFs, topgenes = 10)
                 
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
knitr::kable(tabDEGsTFPubmed, digits = 2, caption = "Table with most studied TF in pubmed related to a specific cancer",row.names = FALSE)
```
