---
title: "Working with TCGAbiolinks package"
author: " Antonio Colaprico, Tiago Chedraoui Silva, Luciano Garofano, 
Catharina Olsen, Davide Garolini, Claudia Cava, Isabella Castiglioni,
Thais Sabedot, Tathiane Malta, Stefano Pagnotta, Michele Ceccarelli
Houtan Noushmehr, Gianluca Bontempi"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        number_sections: false
        toc_depth: 2
        highlight: haddock


references:
- id: ref1
  title: Orchestrating high-throughput genomic analysis with Bioconductor
  author: 
  - family: Huber, Wolfgang and Carey, Vincent J and Gentleman, Robert and Anders, Simon and Carlson, Marc and Carvalho, Benilton S and Bravo, Hector Corrada and Davis, Sean and Gatto, Laurent and Girke, Thomas and others
    given:
  journal: Nature methods
  volume: 12
  number: 2
  pages: 115-121
  issued:
    year: 2015

- id: ref2
  title: GC-content normalization for RNA-Seq data
  author: 
  - family: Risso, Davide and Schwartz, Katja and Sherlock, Gavin and Dudoit, Sandrine
    given:
  journal: BMC bioinformatics
  volume: 12
  number: 1
  pages: 480
  issued:
    year: 2011 
    
- id: ref3
  title: Evaluation of statistical methods for normalization and differential expression in mRNA-Seq experiments
  author: 
  - family: Bullard, James H and Purdom, Elizabeth and Hansen, Kasper D and Dudoit, Sandrine
    given:
  journal: BMC bioinformatics
  volume: 11
  number: 1
  pages: 94
  issued:
    year: 2010 


vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300)
knitr::opts_chunk$set(cache=FALSE)
```

```{r, echo = FALSE,hide=TRUE, message=FALSE,warning=FALSE}
devtools::load_all(".")
```

## Introduction 

Motivation: The Cancer Genome Atlas (TCGA) provides us with an enormous collection of data sets, not only spanning a large number of cancers but also a large number of experimental platforms. Even though the data can be accessed and downloaded from the database, the possibility to analyse these downloaded data directly in one single R package has not yet been available. 

 TCGAbiolinks consists of three parts or levels. Firstly, we provide different options to query and download from TCGA relevant data from all currently platforms and their subsequent pre-processing for commonly used bio-informatics (tools) packages in Bioconductor or CRAN. Secondly, the package allows to integrate different data types and it can be used for different types of analyses dealing with all platforms such as diff.expression, network inference or survival analysis, etc, and then it allows to visualize the obtained results. Thirdly we added a social level where a researcher can found a similar intereset in a bioinformatic community, and allows both to find a validation of results in literature in pubmed and also to retrieve questions and answers from site such as support.bioconductor.org, biostars.org, stackoverflow,etc.

This document describes how to search, download and analyze TCGA data using the
`TCGAbiolinks` package.

## Installation

For the moment the package is in the devel branch of bioconductor repository.
To install use the code below.

```{r, eval = FALSE}
source("http://bioconductor.org/biocLite.R")
useDevel()
biocLite("TCGAbiolinks")
```

# `TCGAquery`: Searching TCGA open-access data  

## `TCGAquery`: Searching TCGA open-access data for download 

 You can easily search TCGA samples using the `TCGAquery` function.
 Using a summary of filters as used in the TCGA portal, the function works
 with the following 
 parameters:

* **tumor** Tumor or list of tumors. The list of tumor is shown in the examples. 
* **platform** Platform or list of tumors. The list of platforms is shown in the examples.
* **samples** List of TCGA barcodes
* **level**  Options: 1,2,3,"mage-tab"
* **center**
* **version** List of Platform/Tumor/Version to be changed


The next subsections will detail each of the search parameters.
Below, we show some search examples:
```{r, eval = FALSE}
query <- TCGAquery(tumor = c("LGG","GBM"), level = 3,
                   platform = c("HumanMethylation450","HumanMethylation27"),
                   samples = c("TCGA-19-4065","TCGA-E1-5322-01A-01D-1467-05"), 
                   version = list(c("HumanMethylation450","LGG",1),
                                  c("HumanMethylation450","GBM",5)))
```

### `TCGAquery`: Searching by tumor
You can filter the search by tumor using the tumor parameter.

```{r, eval = TRUE}
query <- TCGAquery(tumor = "gbm")
```

The list of tumors is show below:
```{r, eval = TRUE, echo = FALSE}
knitr::kable(disease.table, digits = 2,
             caption = "List of tumors",row.names = FALSE)
```

### `TCGAquery`: Searching by level
You can filter the search by level	"1", "2", "3" or "mage-tab"
```{r, eval = TRUE}
query <- TCGAquery(tumor = "gbm", level = 3)
query <- TCGAquery(tumor = "gbm", level = 2)
query <- TCGAquery(tumor = "gbm", level = 1)
query <- TCGAquery(tumor = "gbm", level = "mage-tab")
```

### `TCGAquery`: Searching by platform
You can filter the search by platform using the platform parameter.

```{r, eval = TRUE}
query <- TCGAquery(tumor = "gbm", platform = "IlluminaHiSeq_RNASeqV2")
```

The list of platforms is show below:
```{r, eval = TRUE, echo = FALSE}
knitr::kable(platform.table[,2:4], digits = 2,
             caption = "List of tumors",row.names = FALSE)
```

### `TCGAquery`: Searching by center
You can filter the search by center using the center parameter.

```{r, eval = TRUE}
query <- TCGAquery(tumor = "gbm", center = "mskcc.org")
```

If you don't remember the center or if you have incorrectly typed it. 
It will provide you with all the center names in TCGA. 

The list of centers is show below:
```{r, eval = TRUE, echo = FALSE}
knitr::kable(center.table, digits = 2,
             caption = "List of tumors",row.names = FALSE)
```

### `TCGAquery`: Searching by samples
You can filter the search by samples using the samples parameter.
You can give a list of barcodes or only one barcode. These barcode can be partial barcodes.

```{r, eval = TRUE}
# You can define a list of samples to query and download providing relative TCGA barcodes.
listSamples <- c("TCGA-E9-A1NG-11A-52R-A14M-07","TCGA-BH-A1FC-11A-32R-A13Q-07",
                "TCGA-A7-A13G-11A-51R-A13Q-07","TCGA-BH-A0DK-11A-13R-A089-07",
                "TCGA-E9-A1RH-11A-34R-A169-07","TCGA-BH-A0AU-01A-11R-A12P-07",
                "TCGA-C8-A1HJ-01A-11R-A13Q-07","TCGA-A7-A13D-01A-13R-A12P-07",
                "TCGA-A2-A0CV-01A-31R-A115-07","TCGA-AQ-A0Y5-01A-11R-A14M-07")

# Query all available platforms with a list of barcode 
query <- TCGAquery(samples = listSamples)

# Query with a partial barcode 
query <- TCGAquery(samples = "TCGA-61-1743-01A")
```

## `TCGAquery_Version`: Retrieve versions information of the data in TCGA 

In case the user want to have an overview of the size and number of samples and
date of old versions, you can use the `TCGAquery_Version` function.

The parameters of this function are:

* tumor 
* platform 


For example, the code below queries the version history for the 
IlluminaHiSeq_RNASeqV2 platform .

```{r, eval = FALSE}
library(TCGAbiolinks)

BRCA_RNASeqV2_version <- TCGAquery_Version(tumor = "brca", 
                                     platform = "illuminahiseq_rnaseqv2")

```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
BRCA_RNASeqV2_version <- BRCA_RNASeqV2_version[
    order(BRCA_RNASeqV2_version$SizeMbyte,decreasing=TRUE),]
knitr::kable(BRCA_RNASeqV2_version, digits = 2, 
             caption = "Table with version, number of samples and size (Mbyte) of BRCA IlluminaHiSeq_RNASeqV2 Level 3", 
             row.names = FALSE)

#LatexPrintTableforPresentation(Table = BRCA_RNASeqV2_version,rowsForPage = nrow(BRCA_RNASeqV2_version), TableTitle = "BRCA_RNASeqV2_version", LabelTitle = "BRCA_RNASeqV2_version", withrows = T)

```

### `TCGAquery`: Searching old versions

The results from TCGAquery are always the last one from the TCGA data portal.
As we have a preprocessed table you should always update `TCGAbiolinks` package.
We intent to update the database constantly.

In case you want an old version of the files we have the version parameter 
that should be a list of triple values(platform,tumor,version).
For example the code below will get the LGG and GBM tumor for platform 
HumanMethylation450 but for the LGG/HumanMethylation450, we want the version 5
of the files instead of the latest. This could take some seconds.

```{r, eval = FALSE}
query <- TCGAquery(tumor = c("LGG","GBM"), platform = c("HumanMethylation450"), level = 3, 
                   version = list(c("HumanMethylation450","LGG",1)))
```

## `TCGAquery_clinic` `&` `TCGAquery_clinicFilt`: Working with clinical data.
You can retrive clinical data using the `clinic` function. The parameters of this
function are: 

* cancer ("OV","BRCA","GBM", etc)
* clinical_data_type ("clinical_patient","clinical_drug", etc)

A full list of cancer and clinical data type can be found in the help of the function.

```{r, eval = FALSE}
# Get clinical data
clinical_brca_data <- TCGAquery_clinic("brca","clinical_patient")
clinical_uvm_data_bio <- TCGAquery_clinic("uvm","biospecimen_normal_control")
clinical_brca_data_bio <- TCGAquery_clinic("brca","biospecimen_normal_control")
clinical_brca_data <- TCGAquery_clinic("brca","clinical_patient")
```

Also, some functions to work with clinical data are provided.
For example the function `TCGAquery_clinicFilt` will filter your data, returning the list 
of barcodes that matches all the filter.

The parameters of `TCGAquery_clinicFilt` are:

* **barcode** List of barcodes
* **clinical_patient_data** clinical patient data obtained with clinic function 
Ex: clinical_patient_data <- TCGAquery_clinic("LGG","clinical_patient")
* **HER**  her2 neu immunohistochemistry receptor status: "Positive" or "Negative"
* **gender** "MALE" or "FEMALE"
* **PR**  Progesterone receptor status: "Positive" or "Negative"
* **stage** Pathologic Stage: "stage_IX", "stage_I", "stage_IA", "stage_IB", "stage_IIX",
 "stage_IIA", "stage_IIB", "stage_IIIX","stage_IIIA", "stage_IIIB",
 "stage_IIIC", "stage_IV" -
* **ER** Estrogen receptor status: "Positive" or "Negative"

An example of the function is below:

```{r, eval = FALSE}
bar <- c("TCGA-G9-6378-02A-11R-1789-07", "TCGA-CH-5767-04A-11R-1789-07",  
        "TCGA-G9-6332-60A-11R-1789-07", "TCGA-G9-6336-01A-11R-1789-07",
        "TCGA-G9-6336-11A-11R-1789-07", "TCGA-G9-7336-11A-11R-1789-07",
        "TCGA-G9-7336-04A-11R-1789-07", "TCGA-G9-7336-14A-11R-1789-07",
        "TCGA-G9-7036-04A-11R-1789-07", "TCGA-G9-7036-02A-11R-1789-07",
        "TCGA-G9-7036-11A-11R-1789-07", "TCGA-G9-7036-03A-11R-1789-07",
        "TCGA-G9-7036-10A-11R-1789-07", "TCGA-BH-A1ES-10A-11R-1789-07",
        "TCGA-BH-A1F0-10A-11R-1789-07", "TCGA-BH-A0BZ-02A-11R-1789-07",
        "TCGA-B6-A0WY-04A-11R-1789-07", "TCGA-BH-A1FG-04A-11R-1789-08",
        "TCGA-D8-A1JS-04A-11R-2089-08", "TCGA-AN-A0FN-11A-11R-8789-08",
        "TCGA-AR-A2LQ-12A-11R-8799-08", "TCGA-AR-A2LH-03A-11R-1789-07",
        "TCGA-BH-A1F8-04A-11R-5789-07", "TCGA-AR-A24T-04A-55R-1789-07",
        "TCGA-AO-A0J5-05A-11R-1789-07", "TCGA-BH-A0B4-11A-12R-1789-07",
        "TCGA-B6-A1KN-60A-13R-1789-07", "TCGA-AO-A0J5-01A-11R-1789-07",
        "TCGA-AO-A0J5-01A-11R-1789-07", "TCGA-G9-6336-11A-11R-1789-07",
        "TCGA-G9-6380-11A-11R-1789-07", "TCGA-G9-6380-01A-11R-1789-07",
        "TCGA-G9-6340-01A-11R-1789-07","TCGA-G9-6340-11A-11R-1789-07")

S <- TCGAquery_SampleTypes(bar,"TP")
S2 <- TCGAquery_SampleTypes(bar,"NB")

# Retrieve multiple tissue types  NOT FROM THE SAME PATIENTS
SS <- TCGAquery_SampleTypes(bar,c("TP","NB"))

# Retrieve multiple tissue types  FROM THE SAME PATIENTS
SSS <- TCGAquery_MatchedCoupledSampleTypes(bar,c("NT","TP"))

# Get clinical data
clinical_brca_data <- TCGAquery_clinic("brca","clinical_patient")
female_erpos_herpos <- TCGAquery_clinicFilt(bar,clinical_brca_data, 
                                            HER="Positive", 
                                            gender="FEMALE", 
                                            ER="Positive")
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
bar <- c("TCGA-G9-6378-02A-11R-1789-07", "TCGA-CH-5767-04A-11R-1789-07",  
        "TCGA-G9-6332-60A-11R-1789-07", "TCGA-G9-6336-01A-11R-1789-07",
        "TCGA-G9-6336-11A-11R-1789-07", "TCGA-G9-7336-11A-11R-1789-07",
        "TCGA-G9-7336-04A-11R-1789-07", "TCGA-G9-7336-14A-11R-1789-07",
        "TCGA-G9-7036-04A-11R-1789-07", "TCGA-G9-7036-02A-11R-1789-07",
        "TCGA-G9-7036-11A-11R-1789-07", "TCGA-G9-7036-03A-11R-1789-07",
        "TCGA-G9-7036-10A-11R-1789-07", "TCGA-BH-A1ES-10A-11R-1789-07",
        "TCGA-BH-A1F0-10A-11R-1789-07", "TCGA-BH-A0BZ-02A-11R-1789-07",
        "TCGA-D8-A1JS-04A-11R-2089-08", "TCGA-AN-A0FN-11A-11R-8789-08",
        "TCGA-AR-A2LQ-12A-11R-8799-08", "TCGA-AR-A2LH-03A-11R-1789-07",
        "TCGA-BH-A1F8-04A-11R-5789-07", "TCGA-AR-A24T-04A-55R-1789-07",
        "TCGA-AO-A0J5-05A-11R-1789-07", "TCGA-BH-A0B4-11A-12R-1789-07",
        "TCGA-B6-A1KN-60A-13R-1789-07", "TCGA-AO-A0J5-01A-11R-1789-07",
        "TCGA-AO-A0J5-01A-11R-1789-07", "TCGA-G9-6336-11A-11R-1789-07",
        "TCGA-G9-6380-11A-11R-1789-07", "TCGA-G9-6380-01A-11R-1789-07",
        "TCGA-G9-6340-01A-11R-1789-07","TCGA-G9-6340-11A-11R-1789-07")
female_erpos_herpos <- TCGAquery_clinicFilt(bar, clinBRCA, HER = "Positive", 
                                gender = "FEMALE", ER = "Positive")
print(female_erpos_herpos)
```


## `TCGAquery_subtype`: Working with molecular subtypes data.

The Cancer Genome Atlas (TCGA) Research Network has reported integrated genome-wide 
studies of various diseases. We have added some of the subtypes defined by these
report in our package. The LGG, GBM, STAD, BRCA, READ, COAD and LUAD  tumors 
has data added. These subtypes will be automatically added in the summarizedExperiment
object through TCGAprepare. But you can also use the `TCGAquery_subtype` function
to retrive that information.

```{r, eval = FALSE}
# Check with subtypes from TCGAprepare and update examples
GBM_path_subtypes <- TCGAquery_subtype(tumor = "gbm")

LGG_path_subtypes <- TCGAquery_subtype(tumor = "lgg")

LGG_clinic <- TCGAquery_clinic(cancer = "LGG",clinical_data_type = "clinical_patient")
```

A subset of the lgg subytpe is shown below:

```{r, eval = TRUE, echo = FALSE}
knitr::kable(lgg.subtype[1:10,c(1,2,3,4)], digits = 2, 
             caption = "Table common samples among platforms from TCGAquery",
             row.names = TRUE)
```


## `TCGAquery_integrate`: Summary of the common numbers of patient samples in different platforms 

Some times researches would like to use samples from different platforms 
from the same patient. In order to help the user to have an overview of the number of
samples in commun we created the function `TCGAquery_integrate` that will receive the 
data frame returned from `TCGAquery` and produce a matrix n platforms x n platforms 
with the values of samples in commum.

Some search examples are shown below
```{r, eval = FALSE,echo=TRUE,message=FALSE,warning=FALSE}
query <- TCGAquery(tumor = "brca",level = 3)
matSamples <- TCGAquery_integrate(query)
matSamples[c(1,4,9),c(1,4,9)]
```

The result of the 3 platforms  of `TCGAquery_integrate` result is shown below:

```{r, eval = TRUE, echo = FALSE}
query <- TCGAquery(tumor = "brca",level = 3)
matSamples <- TCGAquery_integrate(query)
knitr::kable(matSamples[c(1,4,9),c(1,4,9)], digits = 2, 
             caption = "Table common samples among platforms from TCGAquery",
             row.names = TRUE)
```




# `TCGAdownload`: Downloading open-access data  
You can easily download data using the `TCGAdownload` function.

The arguments are:

*  **data**	The `TCGAquery` output
*  **path**	location to save the files. Default: "."
*  **type**	Filter the files to download by type
*  **samples**	List of samples to download
*  **force**	Download again if file already exists? Default: FALSE

### `TCGAdownload`: Example of use
```{r, eval = FALSE}
# get all samples from the query and save them in the TCGA folder
# samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results
# samples to normalize later

TCGAdownload(query, path = "data", type = "rsem.genes.results")

TCGAdownload(query, path = "data", type = "rsem.isoforms.normalized_results")


TCGAdownload(query, path = "dataBrca", type = "rsem.genes.results",
             samples = c("TCGA-E9-A1NG-11A-52R-A14M-07",
                         "TCGA-BH-A1FC-11A-32R-A13Q-07")
             )
```

Comment: The function will structure the folders to save the data as:
_Path given by the user/Experiment folder_

### `TCGAdownload`: Table of types available for downloading

*  **RNASeqV2:** junction_quantification,rsem.genes.results,
rsem.isoforms.results, rsem.genes.normalized_results, 
rsem.isoforms.normalized_results, bt.exon_quantification
*  **RNASeq:** exon.quantification,spljxn.quantification, gene.quantification
*  **genome_wide_snp_6:** hg18.seg,hg19.seg,nocnv_hg18.seg,nocnv_hg19.seg

# `TCGAprepare`: Preparing the data
You can easily read the downloaded data using the `TCGAprepare` function.
This function will prepare the data into a 
[SummarizedExperiment](http://www.nature.com/nmeth/journal/v12/n2/abs/nmeth.3252.html) 
[@ref1] object for downstream analysis. 
For the moment this function is working only with data level 3.

The arguments are:

* **query**	Data frame as the one returned from TCGAquery
* **dir**	Directory with the files
* **type**	File to prepare.
* **samples**	List of samples to prepare.
* **save**	Save a rda object with the prepared object? Default: FALSE
* **filename** Name of the rda object that will be saved if `save` is `TRUE`
* **toPackage** Name of the package to prepare the data specific to that package. 
* **summarizedExperiment** Should the output be a SummarizedExperiment object? Default: `TRUE`
* **reannotate** Reannotate genes? Source http://grch37.ensembl.org/. 
Default: `FALSE`. (For the moment only working for methylation data)

In order to add useful information to reasearches we added in the colData of the 
summarizedExperiment the subtypes classification for the LGG and GBM samples 
that can be found in the [TCGA publication section](https://tcga-data.nci.nih.gov/docs/publications/)
We intend to add more tumor types in the future.

Also in the  metadata of the objet we added the parameters used in TCGAprepare,
the query matrix used for preparing, and file information (name,creation time and modification time)
in order to help the user know which samples, versions, and parameters they used.

### Example of use
```{r, eval = FALSE}
# get all samples from the query and save them in the TCGA folder
# samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results
# samples to normalize later
data <- TCGAprepare(query, dir = "data", save = TRUE, filename = "myfile.rda")
```

```{r,eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
library(TCGAbiolinks)
library(SummarizedExperiment)
query <- TCGAquery(tumor = "READ", platform = "IlluminaHiSeq_RNASeqV2", level = 3,
                   samples = c("TCGA-DY-A1DE-01A-11R-A155-07",
                             "TCGA-DY-A0XA-01A-11R-A155-07"))
TCGAdownload(query,samples = c("TCGA-DY-A1DE-01A-11R-A155-07",
                             "TCGA-DY-A0XA-01A-11R-A155-07"),
             type = "rsem.genes.normalized_results",path = "../dataREAD")
dataREAD <- TCGAprepare(query,dir = "../dataREAD", type = "rsem.genes.normalized_results",
                        samples = c("TCGA-DY-A1DE-01A-11R-A155-07",
                             "TCGA-DY-A0XA-01A-11R-A155-07"))
```

As an example, for the platform IlluminaHiSeq_RNASeqV2 we prepared two samples (TCGA-DY-A1DE-01A-11R-A155-07 and TCGA-DY-A0XA-01A-11R-A155-07) for the rsem.genes.normalized_results type. In order to create the object mapped the gene_id 
to the hg19. The genes_id not found are then removed from the final matrix.
The default output is a SummarizedExperiment is shown below. 

```{r, eval = TRUE,message=FALSE,warning=FALSE}
library(TCGAbiolinks)
library(SummarizedExperiment)
head(assay(dataREAD,"normalized_count"))
```

In order to create the SummarizedExperiment object we mapped the rows of the 
experiments into GRanges. In order to map miRNA we used the miRNA from the anotation database TxDb.Hsapiens.UCSC.hg19.knownGene, this will exclude the miRNA from viruses and bacteria. 
In order to map genes, genes alias, we used the biomart hg19 database 
(hsapiens_gene_ensembl from grch37.ensembl.org).

In case you prefere to have the raw data. You can get a data frame without any
modification setting the `summarizedExperiment` to false.

```{r,eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
library(SummarizedExperiment)
library(TCGAbiolinks)
query <- TCGAquery(tumor = "READ", platform = "IlluminaHiSeq_RNASeqV2", level = 3,
                   samples = c("TCGA-DY-A1DE-01A-11R-A155-07",
                             "TCGA-DY-A0XA-01A-11R-A155-07"))
TCGAdownload(query,samples = c("TCGA-DY-A1DE-01A-11R-A155-07",
                             "TCGA-DY-A0XA-01A-11R-A155-07"),
             type = "rsem.genes.normalized_results",path = "../dataREAD")

dataREAD_df <- TCGAprepare(query,dir = "../dataREAD", 
                           type = "rsem.genes.normalized_results",
                        samples = c("TCGA-DY-A1DE-01A-11R-A155-07",
                             "TCGA-DY-A0XA-01A-11R-A155-07"),
                        summarizedExperiment = FALSE)
```

```{r, eval = TRUE}
library(TCGAbiolinks)
class(dataREAD_df)
dim(dataREAD_df)
head(dataREAD_df)
```

### Example of use: Preparing the data with CNV data (Genome_Wide_SNP_6)

You can easily search TCGA samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}

# Define a list of samples to query and download providing relative TCGA barcodes.
samplesList <- c("TCGA-02-0046-10A-01D-0182-01", 
                "TCGA-02-0052-01A-01D-0182-01", 
                "TCGA-02-0033-10A-01D-0182-01", 
                "TCGA-02-0034-01A-01D-0182-01", 
                "TCGA-02-0007-01A-01D-0182-01")

# Query platform Genome_Wide_SNP_6 with a list of barcode 
query <- TCGAquery(tumor = "gbm", level = 3, platform = "Genome_Wide_SNP_6")

# Download a list of barcodes with platform Genome_Wide_SNP_6
TCGAdownload(query, path = "samples")

# Prepare matrix 
GBM_CNV <- TCGAprepare(query, dir = "samples", type = ".hg19.seg.txt")
```

### Table of `types` available for the `TCGAprepare`

*  **RNASeqV2:** junction_quantification,rsem.genes.results,
rsem.isoforms.results, rsem.genes.normalized_results, 
rsem.isoforms.normalized_results, bt.exon_quantification
*  **RNASeq:** exon.quantification,spljxn.quantification, gene.quantification
*  **genome_wide_snp_6:** hg18.seg,hg19.seg,nocnv_hg18.seg,nocnv_hg19.seg

## Preparing the data with parameter - toPackage 
This section will show how to integrate `TCGAbiolinks` with other packages.
Our intention is to provide as many integrations as possible.

The example below shows how to use `TCGAbiolinks` with `ELMER` package 
(expression/methylation analysis). The TCGAprepare for the DNA methylation data 
will Removing probes with NA values in more than 0.80% samples and remove the 
anottation data, fot the expression data it will take the log2(expression + 1)
of the expression matrix in order to To linearize the relation between 
DNA methylation and expressionm also it will prepare the rownames as the 
specified by the package.

```{r, eval = FALSE}
############## Get tumor samples with TCGAbiolinks
library(TCGAbiolinks)
query <- TCGAquery(tumor = "GBM",level = 3, platform = "HumanMethylation450")
# This function will take a lot of time depens on internet connection
TCGAdownload(query,path = "TCGA/450k")
met <- TCGAprepare(query,dir = "TCGA/450k",
                   save = TRUE,
                   filename = "met.rda",
                   toPackage = "ELMER")

query.rna <- TCGAquery(tumor="GBM",level=3, platform="IlluminaHiSeq_RNASeqV2")
TCGAdownload(query.rna,path="TCGA/rna",type = "rsem.genes.normalized_results")
exp <- TCGAprepare(query.rna, dir="TCGA/rna", save = TRUE, 
                    filename = "exp.rda",toPackage = "ELMER")

############# To EMLER
library(ELMER)

########## gene annotation
geneAnnot <- txs()
geneAnnot$GENEID <- paste0("ID",geneAnnot$GENEID)
geneInfo <- promoters(geneAnnot,upstream = 0, downstream = 0)
############ probe
probe <- get.feature.probe()

mee.gbm.glial.with.exp <- fetch.mee(meth = gbm.glial.m,
                                exp = exp,
                                probeInfo = probe,
                                TCGA = TRUE,
                                geneInfo = geneInfo)
```




# `TCGAanalyze`: Analyze data from TCGA. 
You can easily analyze data using following functions:

## `TCGAanalyze_Preprocessing` Preprocessing of Gene Expression data (IlluminaHiSeq_RNASeqV2).

You can easily search TCGA samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}

# You can define a list of samples to query and download providing relative TCGA barcodes.

listSamples <- c("TCGA-E9-A1NG-11A-52R-A14M-07","TCGA-BH-A1FC-11A-32R-A13Q-07",
                "TCGA-A7-A13G-11A-51R-A13Q-07","TCGA-BH-A0DK-11A-13R-A089-07",
                "TCGA-E9-A1RH-11A-34R-A169-07","TCGA-BH-A0AU-01A-11R-A12P-07",
                "TCGA-C8-A1HJ-01A-11R-A13Q-07","TCGA-A7-A13D-01A-13R-A12P-07",
                "TCGA-A2-A0CV-01A-31R-A115-07","TCGA-AQ-A0Y5-01A-11R-A14M-07")

# Query platform IlluminaHiSeq_RNASeqV2 with a list of barcode 
query <- TCGAquery(tumor = "brca", samples = listSamples, 
  platform = "IlluminaHiSeq_RNASeqV2", level = "3")

# Download a list of barcodes with platform IlluminaHiSeq_RNASeqV2
TCGAdownload(query, path = "../dataBrca", type = "rsem.genes.results",samples = listSamples)

# Prepare expression matrix with gene id in rows and samples (barcode) in columns
# rsem.genes.results as values
BRCARnaseq_assay <- TCGAprepare(query,"../dataBrca",type = "rsem.genes.results")

BRCAMatrix <- assay(BRCARnaseq_assay,"raw_counts")

# For gene expression if you need to see a boxplot correlation and AAIC plot 
# to define outliers you can run
BRCARnaseq_CorOutliers <- TCGAanalyze_Preprocessing(BRCARnaseq_assay)
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE,size = 8}
library(TCGAbiolinks)
dataGE <- dataBRCA[sample(rownames(dataBRCA),10),sample(colnames(dataBRCA),7)]

knitr::kable(dataGE[1:10,2:3], digits = 2, 
            caption = "Example of a matrix of gene expression (10 genes in rows and 2 samples in columns)",
            row.names = TRUE)
```

The result from TCGAanalyze_Preprocessing is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("PreprocessingOutput.png")
grid.raster(img)
```


## `TCGAanalyze_DEA` `&` `TCGAanalyze_LevelTab` Differential expression analysis (DEA)

Perform DEA (Differential expression analysis) to identify differentially expressed genes (DEGs) using the `TCGAanalyze_DEA` function. 

`TCGAanalyze_DEA` performs DEA using following functions from R \Biocpkg{edgeR}:

1. edgeR::DGEList converts the count matrix into an edgeR object.
2. edgeR::estimateCommonDisp each gene gets assigned the same dispersion estimate.
3. edgeR::exactTest performs pair-wise tests for differential expression between two groups.
4. edgeR::topTags takes the output from exactTest(), adjusts the raw p-values using the False Discovery Rate (FDR) correction, and returns the top differentially expressed genes.

This function receives as parameters:

* **mat1** The matrix of the first group (in the example group 1 is the normal samples), 
* **mat2** The matrix of the  second group (in the example group 2 is tumor samples)
* **Cond1type** Label for group 1
* **Cond1type** Label for group 2

After, we filter the output of dataDEGs by abs(LogFC) >=1, and uses the
`TCGAanalyze_LevelTab` function to create a table with DEGs (differentially expressed genes), log Fold Change (FC), false discovery rate (FDR), the gene expression level for samples in Cond1type, and Cond2type, and Delta value (the difference of gene expression between the two conditions multiplied logFC).

```{r, eval = TRUE}
# Downstream analysis using gene expression data  
# TCGA samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results
# save(dataBRCA, geneInfo , file = "dataGeneExpression.rda")
library(TCGAbiolinks)

# normalization of genes
dataNorm <- TCGAanalyze_Normalization(tabDF = dataBRCA,geneInfo =  geneInfo)

# quantile filter of genes
dataFilt <- TCGAanalyze_Filtering(tabDF = dataNorm,
                                  method = "quantile", 
                                  qnt.cut =  0.25)

# selection of normal samples "NT"
samplesNT <- TCGAquery_SampleTypes(barcode = colnames(dataFilt),
                                   typesample = c("NT"))

# selection of tumor samples "TP"
samplesTP <- TCGAquery_SampleTypes(barcode = colnames(dataFilt), 
                                   typesample = c("TP"))

# Diff.expr.analysis (DEA)
dataDEGs <- TCGAanalyze_DEA(mat1 = dataFilt[,samplesNT],
                            mat2 = dataFilt[,samplesTP],
                            Cond1type = "Normal",
                            Cond2type = "Tumor",
                            fdr.cut = 0.01 ,
                            logFC.cut = 1,
                            method = "glmLRT")

# DEGs table with expression values in normal and tumor samples
dataDEGsFiltLevel <- TCGAanalyze_LevelTab(dataDEGs,"Tumor","Normal",
                                dataFilt[,samplesTP],dataFilt[,samplesNT])

```
The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
dataDEGsFiltLevel$FDR <- format(dataDEGsFiltLevel$FDR, scientific = TRUE)
knitr::kable(dataDEGsFiltLevel[1:10,], digits = 2,
                caption = "Table DEGs after DEA",row.names = FALSE)
```

## `TCGAanalyze_EAcomplete & TCGAvisualize_EAbarplot`: Enrichment Analysis

Researchers, in order to better understand the underlying biological processes, 
often want to retrieve a functional profile of a set of genes that might 
have an important role. This can be done by performing an enrichment analysis. 

We will perform an enrichment analysis on gene sets using the `TCGAanalyze_EAcomplete` 
function. 
Given a set of genes that are up-regulated under certain conditions, 
an enrichment analysis will find  identify classes of genes or proteins that 
are over-represented  using annotations for that gene set.


To view the results you can use the `TCGAvisualize_EAbarplot` function as shown below.

```{r, eval = FALSE}
library(TCGAbiolinks)
# Enrichment Analysis EA
# Gene Ontology (GO) and Pathway enrichment by DEGs list
Genelist <- rownames(dataDEGsFiltLevel)

system.time(ansEA <- TCGAanalyze_EAcomplete(TFname="DEA genes Normal Vs Tumor",Genelist))

# Enrichment Analysis EA (TCGAVisualize)
# Gene Ontology (GO) and Pathway enrichment barPlot

TCGAvisualize_EAbarplot(tf = rownames(ansEA$ResBP), 
            GOBPTab = ansEA$ResBP,
            GOCCTab = ansEA$ResCC,
            GOMFTab = ansEA$ResMF,
            PathTab = ansEA$ResPat,
            nRGTab = Genelist, 
            nBar = 10)

```
The result is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("EAplot.png")
grid.raster(img)
```

## `TCGAanalyze_survival` Survival Analysis: Cox Regression and dnet package
When analyzing survival times, different problems come up than the ones dis-
cussed so far. One question is how do we deal with subjects dropping out of a
study. For example, assume that we test a new cancer drug. While some subjects
die, others may believe that the new drug is not effective, and decide to drop out
of the study before the study is finished. A similar problem would be faced when
we investigate how long a machine lasts before it breaks down.
 
Using the clinical data, it is possible to create a survival plot with the 
function `TCGAanalyze_survival` as follows:

```{r, eval = FALSE}
clin.gbm <- TCGAquery_clinic("gbm", "clinical_patient")
clin.lgg <- TCGAquery_clinic("lgg", "clinical_patient")

TCGAanalyze_survival(plyr::rbind.fill(clin.lgg,clin.gbm),
                     "radiation_therapy",
                     main = "TCGA Set\nLGG and GBM",height = 10, width=10)
```

The arguments of `TCGAanalyze_survival` are:

*  **clinical_patient** TCGA Clinical patient with the information days_to_death
*  **clusterCol** Column with groups to plot. This is a mandatory field, 
the caption will be based in this column
*  **legend** Legend title of the figure
*  **cutoff** xlim This parameter will be a limit in the x-axis. 
That means, that patients with days_to_deth > cutoff will be set to Alive.
*  **main**	 main title of the plot
*  **ylab**	y-axis text of the plot
*  **xlab** x-axis text of the plot
*  **filename**	 The name of the pdf file
*  **color** Define the colors of the lines.

 The result is shown below:
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("survival.png")
grid.raster(img)
```

 
```{r, eval = FALSE}
library(TCGAbiolinks)
# Survival Analysis SA

clinical_patient_Cancer <- TCGAquery_clinic("brca","clinical_patient")
dataBRCAcomplete <- log2(BRCA_rnaseqv2)

tokenStop<- 1

tabSurvKMcomplete <- NULL

for( i in 1: round(nrow(dataBRCAcomplete)/100)){
message( paste( i, "of ", round(nrow(dataBRCAcomplete)/100)))
tokenStart <- tokenStop
tokenStop <-100*i
tabSurvKM<-TCGAanalyze_SurvivalKM(clinical_patient_Cancer,dataBRCAcomplete,
                                Genelist = rownames(dataBRCAcomplete)[tokenStart:tokenStop],
                                        Survresult = F,ThreshTop=0.67,ThreshDown=0.33)

tabSurvKMcomplete <- rbind(tabSurvKMcomplete,tabSurvKM)
}

tabSurvKMcomplete <- tabSurvKMcomplete[tabSurvKMcomplete$pvalue < 0.01,]
tabSurvKMcomplete <- tabSurvKMcomplete[!duplicated(tabSurvKMcomplete$mRNA),]
rownames(tabSurvKMcomplete) <-tabSurvKMcomplete$mRNA
tabSurvKMcomplete <- tabSurvKMcomplete[,-1]
tabSurvKMcomplete <- tabSurvKMcomplete[order(tabSurvKMcomplete$pvalue, decreasing=F),]
    
tabSurvKMcompleteDEGs <- tabSurvKMcomplete[rownames(tabSurvKMcomplete) %in% dataDEGsFiltLevel$mRNA,]
```
The result is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
tabSurvKMcompleteDEGs$pvalue <- format(tabSurvKMcompleteDEGs$pvalue, scientific = TRUE)
knitr::kable(tabSurvKMcompleteDEGs[1:10,1:4], 
            digits = 2,
            caption = "Table KM-survival genes after SA",
            row.names = TRUE)
knitr::kable(tabSurvKMcompleteDEGs[1:10,5:7], 
            digits = 2,
            row.names = TRUE)
```


## `TCGAanalyze_DMR`: Differentially methylated regions Analysis

We will search for differentially methylated CpG sites using the `TCGAanalyze_DMR`
function. In order to find these regions we use the beta-values 
(methylation values ranging from 0.0 to 1.0) to compare two groups. 

Firstly, it calculates the difference between the mean DNA methylation of each group
for each probes. 

Secondly, it calculates the p-value using the wilcoxon test 
adjusting by the Benjamini-Hochberg method. The default parameters was set to
 require a minimum absolute beta-values difference of 0.2 and a p-value adjusted
 of < 0.01. 

After these analysis, we save a volcano plot (x-axis:diff mean methylation,
y-axis: significance) that will help the user identify the differentially
methylated CpG sites and return the object with the calculus in the rowRanges.

The arguments of volcanoPlot are:

* **data** SummarizedExperiment obtained from the TCGAPrepare
* **groupCol**  Columns with the groups inside the SummarizedExperiment
  object. (This will be obtained by the function colData(data))
* **group1 ** In case our object has more than 2 groups, you should set
 the name of the group
* **group2 ** In case our object has more than 2 groups, you should set
 the name of the group
* **filename ** pdf filename. Default: volcano.pdf
* **legend ** Legend title
* **color **vector of colors to be used in graph
* **title **main title. If not specified it will be
 "Volcano plot (group1 vs group2)
* **ylab **y axis text
* **xlab **x axis text
* **xlim **x limits to cut image
* **ylim **y limits to cut image
* **label **vector of labels to be used in the figure.
 Example: c("1" = "Not Significant", "2" = "Hypermethylated in group1",
 "3" = "Hypomethylated in group1"))
* **p.cut** p values threshold. *Default: 0.01*
* **diffmean.cut** diffmean threshold. *Default: 0.2*
* **adj.method** Adjusted method for the p-value calculation
* **paired** Wilcoxon paired parameter. *Default: FALSE*
* **overwrite** Overwrite the pvalues and diffmean values if already in the object
for both groups? *Default: FALSE*

```{r, eval = FALSE}
data <- TCGAanalyze_DMR(data, groupCol = "cluster.meth",subgroupCol = "disease",
                              group.legend  = "Groups", subgroup.legend = "Tumor",
                              print.pvalue = TRUE)
```

The output will be a plot such as the figure below. The green dots are the 
probes that are hypomethylated in group 2 compared to group 1, 
while the red dots are the hypermethylated probes in group 2 compared to group 1


```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure5met.png")
grid.raster(img)
```


Also, the `TCGAanalyze_DMR` 
function will save the plot as pdf and return the same SummarizedExperiment
that was given as input with the values of p-value,
p-value adjusted, diffmean and the group it belongs in the graph 
(non significant, hypomethylated, hypermethylated) in the rowRanges. 
The collumns will be (where group1 and group2 are the names of the groups): 

* diffmean.group1.group2 (mean.group2 - mean.group1) 
* diffmean.group2.group1 (mean.group1 - mean.group2)
* p.value.group1.group2
* p.value.adj.group1.group2
* status.group1.group2 (Status of probes in group2 in relation to group1) 
* status.group2.group1 (Status of probes in group1 in relation to group2) 

This values can be view/acessed using the `rowRanges` 
acessesor (`rowRanges(data)`).

**Observation:** Calling the same function again, with the same arguments 
will only plot the results,  as it was already calculated. 
With you want to have them recalculated, please set
`overwrite` to `TRUE` or remove the calculated collumns.



# `TCGAvisualize`: Visualize results from analysis functions with TCGA's data. 
You can easily visualize results from soome following functions:

## `TCGAvisualize_PCA`: Principal Component Analysis plot for differentially expressed genes

In order to understand better our genes, we can perform a PCA to reduce the
number of dimensions of our gene set. The function `TCGAvisualize_PCA` will plot
the PCA for different groups.

The parameters of this function are:

* **dataFilt** The expression matrix after normalization and quantile filter
* **dataDEGsFiltLevel** The TCGAanalyze_LevelTab output
* **ntopgenes** number of DEGs genes to plot in PCA

```{r, eval = FALSE}
# normalization of genes
dataNorm <- TCGAbiolinks::TCGAanalyze_Normalization(dataBRCA, geneInfo)

# quantile filter of genes
dataFilt <- TCGAanalyze_Filtering(tabDF = dataNorm,
                                  method = "quantile", 
                                  qnt.cut =  0.25)

# Principal Component Analysis plot for ntop selected DEGs
TCGAvisualize_PCA(dataFilt,dataDEGsFiltLevel, ntopgenes = 200)
```

The result is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}
library(png)
library(grid)
img <- readPNG("PCAtop200DEGs.png")
grid.raster(img)
```

## `TCGAvisualize_SurvivalCoxNET` Survival Analysis: Cox Regression and dnet package

TCGAvisualize_SurvivalCoxNET can help an user to identify a group of survival genes that are significant from univariate Kaplan Meier Analysis and also for Cox Regression.
It shows in the end a network build with community of genes with similar range of pvalues from
Cox regression (same color) and that interaction among those genes is already validated in
literatures using the STRING database (version 9.1).


```{r, eval = FALSE}
library(TCGAbiolinks)
# Survival Analysis SA

clinical_patient_Cancer <- TCGAquery_clinic("brca","clinical_patient")
dataBRCAcomplete <- log2(BRCA_rnaseqv2)

tokenStop<- 1

tabSurvKMcomplete <- NULL

for( i in 1: round(nrow(dataBRCAcomplete)/100)){
message( paste( i, "of ", round(nrow(dataBRCAcomplete)/100)))
tokenStart <- tokenStop
tokenStop <-100*i
tabSurvKM<-TCGAanalyze_SurvivalKM(clinical_patient_Cancer,
                                dataBRCAcomplete,
                                Genelist = rownames(dataBRCAcomplete)[tokenStart:tokenStop],
                                Survresult = F,ThreshTop=0.67,ThreshDown=0.33)

tabSurvKMcomplete <- rbind(tabSurvKMcomplete,tabSurvKM)
}

tabSurvKMcomplete <- tabSurvKMcomplete[tabSurvKMcomplete$pvalue < 0.01,]
tabSurvKMcomplete <- tabSurvKMcomplete[!duplicated(tabSurvKMcomplete$mRNA),]
rownames(tabSurvKMcomplete) <-tabSurvKMcomplete$mRNA
tabSurvKMcomplete <- tabSurvKMcomplete[,-1]
tabSurvKMcomplete <- tabSurvKMcomplete[order(tabSurvKMcomplete$pvalue, decreasing=F),]

tabSurvKMcompleteDEGs <- tabSurvKMcomplete[rownames(tabSurvKMcomplete) %in% dataDEGsFiltLevel$mRNA,]

tflist <- EAGenes[EAGenes$Family == "transcription regulator","Gene"]
tabSurvKMcomplete_onlyTF <- tabSurvKMcomplete[rownames(tabSurvKMcomplete) %in% tflist,]

TabCoxNet <- TCGAvisualize_SurvivalCoxNET(clinical_patient_Cancer,dataBRCAcomplete,
                            Genelist = rownames(tabSurvKMcompleteDEGs),
                            scoreConfidence = 700,titlePlot = "TCGAvisualize_SurvivalCoxNET Example")
```
In particular the survival analysis with kaplan meier and cox regression allow user to reduce the feature / number of genes significant for survival. And using 'dnet' pipeline with 'TCGAvisualize_SurvivalCoxNET' function the user can further filter those genes according some already validated interaction according STRING database.
This is important because the user can have an idea about the biology inside the 
survival discrimination and further investigate in a sub-group of genes that are working in as synergistic effect influencing the risk of survival.
In the following picture the user can see some community of genes with same color and survival pvalues.

The result is shown below:
```{r, fig.width=6, fig.height=4, echo=FALSE, fig.align="center"}

library(png)
library(grid)
img <- readPNG("SurvivalCoxNETOutput.png")
grid.raster(img)
```

## `TCGAvisualize_meanMethylation`: Sample Mean DNA Methylation Analysis
Using the data and calculating the mean DNA methylation per group, it is possible 
to create a mean DNA methylation boxplot with the function `TCGAvisualize_meanMethylation`
as follows:

```{r, eval = FALSE}
TCGAvisualize_meanMethylation(data,"group")
```
The arguments of `TCGAvisualize_meanMethylation` are:

* **data** SummarizedExperiment object obtained from `TCGAPrepare`
* **groupCol** Columns in colData(data) that defines the groups. If no
 columns defined a columns called "Patients" will be used
* **subgroupCol** Columns in colData(data) that defines the subgroups.
* **shapes** Shape vector of the subgroups. It must have the size of the levels
of the subgroups. Example: shapes = c(21,23) if for two levels
* **filename** The name of the pdf that will be saved
* **subgroup.legend** Name of the subgroup legend. **DEFAULT: subgroupCol**
* **group.legend** Name of the group legend. **DEFAULT: groupCol**
* **color** vector of colors to be used in graph
* **title** main title in the plot
* **ylab** y axis text in the plot
* **print.pvalue** Print p-value for two groups in the plot
* **xlab** x axis text in the plot
* **labels** Labels of the groups

The result is shown below:
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("meanmet.png")
grid.raster(img)
```


## `TCGAvisualize_starburst`: Analyzing expression and methylation together 

 The starburst plot is proposed to combine information from two volcano plots, 
 and is applied for a study of DNA methylation and gene expression. In order to
 reproduce this plot, we will use the `TCGAvisualize_starburst` function.
 
The function creates Starburst plot for comparison of DNA methylation
and gene expression. The log10 (FDR-corrected P value) is plotted for beta 
value for DNA methylation (x axis) and gene expression (y axis) for each gene. 
The black dashed line shows the FDR-adjusted P value of 0.01.

The parameters of this function are:

* **met** SummarizedExperiment with methylation data obtained from the 
`TCGAprepare` and processed by `TCGAanalyze_DMR` function. 
Expected colData columns: diffmean and p.value.adj 
* **exp** Matrix with expression data obtained from the `TCGAanalyze_DEA` function. 
Expected colData columns: logFC, FDR
* **filename**  pdf filename
* **legend**  legend title
* **color**  vector of colors to be used in graph
* **label**  vector of labels to be used in graph
* **title**  main title
* **ylab**  y axis text
* **xlab**  x axis text
* **xlim**  x limits to cut image
* **ylim**  y limits to cut image
* **p.cut**  p value cut-off
* **group1**  The name of the group 1  Obs: Column p.value.adj.group1.group2 should exist
* **group2**  The name of the group 2. Obs: Column p.value.adj.group1.group2 should exist
* **exp.p.cut** expression p value cut-off
* **met.p.cut**	methylation p value cut-off
* **diffmean.cut**	If set, the probes with diffmean higher than methylation cut-off will be highlighted in the plot. And the data frame return will be subseted.
* **logFC.cut** If set, the probes with expression fold change higher than methylation cut-off will be highlighted in the plot. And the data frame return will be subseted.


```{r, eval = FALSE}
starburst <- TCGAvisualize_starburst(coad.SummarizeExperiment, 
                                     different.experssion.analysis.data,
                                     group1 = "CIMP.H",
                                     group2 = "CIMP.L",
                                     met.p.cut = 10^-5, 
                                     exp.p.cut=10^-5,
                                     names = TRUE)
```

As result the function will a plot the figure below and return a matrix with 
The Gene_symbol and it status in relation to expression(up regulated/down regulated) and 
methylation (Hyper/Hypo methylated). The case study 3, shows the complete pipeline
for creating this figure.


```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure5star.png")
grid.raster(img)
```


# `TCGAquery_investigate`: Find most studied TFs in pubmed 

Find most studied TFs in pubmed related to a specific cancer, disease, or tissue
 
```{r, eval = FALSE}
# First perform DEGs with TCGAanalyze
# See previous section 
library(TCGAbiolinks)

# Select only transcription factors (TFs) from DEGs
TFs <- EAGenes[EAGenes$Family =="transcription regulator",]
TFs_inDEGs <- intersect(TFs$Gene, dataDEGsFiltLevel$mRNA )
dataDEGsFiltLevelTFs <- dataDEGsFiltLevel[TFs_inDEGs,]

# Order table DEGs TFs according to Delta decrease
dataDEGsFiltLevelTFs <- dataDEGsFiltLevelTFs[order(dataDEGsFiltLevelTFs$Delta,decreasing = TRUE),]

# Find Pubmed of TF studied related to cancer
tabDEGsTFPubmed <- TCGAquery_investigate("breast", dataDEGsFiltLevelTFs, topgenes = 10)
                 
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
library(stringr)
tabDEGsTFPubmed$PMID <- str_sub(tabDEGsTFPubmed$PMID,0,30)
knitr::kable(tabDEGsTFPubmed, digits = 2, caption = "Table with most studied TF in pubmed related to a specific cancer",row.names = FALSE)

#LatexPrintTableforPresentation(Table = tabDEGsTFPubmed,rowsForPage = nrow(tabDEGsTFPubmed), TableTitle = "tabDEGsTFPubmed", LabelTitle = "tabDEGsTFPubmed", withrows = T)
```

# `TCGAquery_Social`: Searching questions,answers and literature
The `TCGAquery_Social` function has two type of searches, one that searches for
most downloaded packages in CRAN or BioConductor and one that searches the most related question in biostar. 

## `TCGAquery_Social` with BioConductor
Find most downloaded packages in CRAN or BioConductor
 
```{r, eval = FALSE}
library(TCGAbiolinks)

# Define a list of package to find number of downloads
listPackage <-c("limma","edgeR","survcomp")

tabPackage <- TCGAquery_Social(siteToFind ="bioconductor.org",listPackage)


# define a keyword to find in support.bioconductor.org returing a table with suggested packages
tabPackageKey <- TCGAquery_Social(siteToFind ="support.bioconductor.org" ,KeyInfo = "tcga")
```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
listPackage <- c("limma","edgeR","survcomp")
tabPackage <- TCGAquery_Social(siteToFind = "bioconductor.org", listPackage)

knitr::kable(head(tabPackage), digits = 2, caption = "Table with number of downloads about a list of packages",row.names = FALSE)

knitr::kable(tabPackage2[c(1,4,5,7),], digits = 2, caption = "Find most related question in support.bioconductor.org with keyword = tcga",row.names = FALSE)

```

## `TCGAquery_Social` with Biostar
Find most related question in biostar.
 
```{r, eval = FALSE}
library(TCGAbiolinks)

# Find most related question in biostar with TCGA
tabPackage1 <- TCGAquery_Social(siteToFind ="biostars.org",KeyInfo = "TCGA")

# Find most related question in biostar with package
tabPackage2 <- TCGAquery_Social(siteToFind ="biostars.org",KeyInfo = "package")

```

The result is shown below:

```{r, eval = TRUE, echo = FALSE}
library(TCGAbiolinks)
# Find most related question in biostar with TCGA
#tabPackage <- TCGAquery_Social(siteToFind ="biostars.org",KeyInfo = "package")

knitr::kable(head(tabPackage1), digits = 2, caption = "Find most related question in biostar with TCGA",row.names = FALSE)

knitr::kable(tabPackage2[c(1,5,7),], digits = 2, caption = "Find most related question in biostar with package",row.names = FALSE)
```

## TCGA Downstream Analysis some workflows and pipelines.

# Downstream Analysis n.1 IlluminaHiSeq_RNASeqV2 data
After preparing the gene expression from TCGA data using the `TCGAprepare`
function, you can do a normalization of genes using the 
function `TCGAanalyze_Normalization`, do a quantile filter of genes with
the `TCGAanalyze_Filtering` function. 

TCGAanalyze_Normalization allows user to normalize mRNA transcripts and miRNA, 
using R \Biocpkg{EDASeq} package. Normalization for RNA-Seq Numerical and graphical summaries of RNA-Seq read data. Within-lane normalization procedures to adjust for GC-content effect (or other gene-level effects) on read counts: loess robust local regression, global-scaling, and full-quantile normalization [@ref2]. Between-lane normalization procedures to adjust for distributional differences between lanes (e.g., sequencing depth): global-scaling and full-quantile normalization [@ref3].

For istance returns all mRNA or miRNA with mean across all samples, higher than the threshold defined quantile mean across all samples.


Also, in order to classify your samples (barcode) 
you can use the `TCGAquery_SampleTypes` function, the typeSample "NT" will return 
the  "Solid Tissue Normal" samples, while the typeSample "TP" will return 
"Primary Solid Tumor"" samples.
 
```{r, eval = TRUE}
# Downstream analysis using gene expression data  
# TCGA samples from IlluminaHiSeq_RNASeqV2 with type rsem.genes.results

library(TCGAbiolinks)

# dataBRCA in TCGAbiolinks package is a table from TCGA BRCA [10 samples] and comes from 
# BRCAMatrix <- TCGAprepare(query,"dataBrca") from above example
# dataBRCA <- BRCAMatrix

# normalization of genes
dataNorm <- TCGAbiolinks::TCGAanalyze_Normalization(dataBRCA, geneInfo)

# quantile filter of genes
dataFilt <- TCGAanalyze_Filtering(tabDF = dataNorm,
                                  method = "quantile", 
                                  qnt.cut =  0.25)

# selection of normal samples "NT"
samplesNT <- TCGAquery_SampleTypes(colnames(dataFilt), typesample = c("NT"))

# selection of tumor samples "TP"
samplesTP <- TCGAquery_SampleTypes(colnames(dataFilt), typesample = c("TP"))


```

## Downstream Analysis n.2 IlluminaHiSeq_RNASeq data

You can easily search TCGA samples, download and prepare a matrix of gene expression.
```{r, eval = FALSE}
# Query platform IlluminaHiSeq_RNASeq withot a list of barcode 
query <- TCGAquery(tumor = "brca", platform = "IlluminaHiSeq_RNASeq", level = "3")

# You can define a list of samples to query and download providing relative TCGA barcodes.
listSamples <- TCGAquery_samplesfilter(query)

# Download only first 5 samples for test.

TCGAdownload(query, path = "dataBrca", type = "gene.quantification",
             samples = listSamples$IlluminaHiSeq_RNASeq[1:5])

# Prepare expression matrix with gene id in rows and samples (barcode) in columns
# rsem.genes.results as values
BRCAMatrix <- TCGAprepare(query,"dataBrca",type = "gene.quantification")
```


# TCGA Downstream Analysis: Case Studies

## Introduction 

This vignette shows a complete workflow of the TCGAbiolinks package. 
The code is divided in 4 case study:

* 1. Expression pipeline (BRCA)
* 2. Expression pipeline (GBM)
* 3. Methylation pipeline - LGG/OV
* 4. Elmer pipeline - KIRC

## Parameters definition
```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
PlatformCancer <- "IlluminaHiSeq_RNASeqV2"
dataType <- "rsem.genes.results"
pathGBM<- "../dataGBM"
pathLGG <- "../dataLGG"

library(BiocInstaller)
useDevel()  # we need Devel for SummarizedExperiment package
library(SummarizedExperiment)
library(TCGAbiolinks)
```

## Case study n. 1: Pan Cancer downstream analysis BRCA

```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
library(TCGAbiolinks)

cancer <- "BRCA"
PlatformCancer <- "IlluminaHiSeq_RNASeqV2"
dataType <- "rsem.genes.results"
pathCancer <- paste0("../data",cancer)

# Result....Function.....parameters...p1...pn..................................time execution

datQuery <- TCGAquery(tumor = cancer, platform = PlatformCancer, level = "3")  # time = 0.093s
lsSample <- TCGAquery_samplesfilter(query = datQuery)
dataSubt <- TCGAquery_subtype(tumor = cancer)
dataSmTP <- TCGAquery_SampleTypes(barcode = lsSample$IlluminaHiSeq_RNASeqV2,
                                  typesample = "TP")
dataSmTN <- TCGAquery_SampleTypes(barcode = lsSample$IlluminaHiSeq_RNASeqV2,
                                  typesample ="NT")
dataClin <- TCGAquery_clinic(cancer = cancer,
                             clinical_data_type = "clinical_patient")

            TCGAdownload(data = datQuery, path = pathCancer, type = dataType,
                         samples =c(dataSmTP,dataSmTN))

dataAssy <- TCGAprepare(query = datQuery, 
                        dir = pathCancer, 
                        type = dataType, 
                        save = TRUE, 
                        summarizedExperiment = TRUE, 
                        samples = c(dataSmTP,dataSmTN),
                        filename = paste0(cancer,"_",PlatformCancer,".rda"))

dataPrep <- TCGAanalyze_Preprocessing(object = dataAssy,cor.cut = 0.6)
dataNorm <- TCGAanalyze_Normalization(tabDF = dataPrep,
                                      geneInfo = geneInfo,
                                      method = "gcContent")
dataFilt <- TCGAanalyze_Filtering(tabDF = dataNorm,
                                  method = "quantile", 
                                  qnt.cut =  0.25)
dataDEGs <- TCGAanalyze_DEA(mat1 = dataFilt[,dataSmTN],
                            mat2 = dataFilt[,dataSmTP],
                            Cond1type = "Normal",
                            Cond2type = "Tumor",
                            fdr.cut = 0.01 ,
                            logFC.cut = 1,
                            method = "glmLRT")

dataSurv<-TCGAanalyze_SurvivalKM(clinical_patient = clinical_patient_Cancer,
                                 dataGE = dataFilt,
                                 
dataSurv<-TCGAanalyze_SurvivalKM(clinical_patient = dataClin,
                                dataGE = dataFilt,
                                Genelist = rownames(dataDEGs)[1:50],
                                Survresult = FALSE,
                                ThreshTop = 0.67,
                                ThreshDown = 0.33)
```


## Case study n. 2: Pan Cancer downstream analysis LGG

```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
library(TCGAbiolinks)
cancer <- "LGG"
PlatformCancer <- "IlluminaHiSeq_RNASeqV2"
dataType <- "rsem.genes.results"
pathCancer <- paste0("../data",cancer)

# Result....Function.....parameters...p1...pn..................................time execution

datQuery <- TCGAquery(tumor = cancer, platform = PlatformCancer, level = "3")  # time = 0.093s
lsSample <- TCGAquery_samplesfilter(query = datQuery)
dataSubt <- TCGAquery_subtype(tumor = cancer)
dataSmTP <- TCGAquery_SampleTypes(barcode = lsSample$IlluminaHiSeq_RNASeqV2,
                                  typesample = "TP")
dataClin <- TCGAquery_clinic(cancer = cancer,
                             clinical_data_type = "clinical_patient")

            TCGAdownload(data = datQuery, path = pathCancer, type = dataType,
                         samples = dataSmTP )

dataAssy <- TCGAprepare(query = datQuery, 
                        dir = pathCancer, 
                        type = dataType, 
                        save = TRUE, 
                        summarizedExperiment = TRUE, 
                        samples = dataSmTP,
                        filename = paste0(cancer,"_",PlatformCancer,".rda"))

dataPrep <- TCGAanalyze_Preprocessing(object = dataAssy,cor.cut = 0.6)        # time = 13.028s
dataNorm <- TCGAanalyze_Normalization(tabDF = dataPrep,
                                      geneInfo = geneInfo,
                                      method = "gcContent")                  # time = 165.577s 

datFilt1 <- TCGAanalyze_Filtering(tabDF = dataNorm,method = "varFilter")
datFilt2 <- TCGAanalyze_Filtering(tabDF = datFilt1,method = "filter1")
datFilt <- TCGAanalyze_Filtering(tabDF = datFilt2,method = "filter2")

data_Hc1 <- TCGAanalyze_Clustering(tabDF = datFilt,method = "hclust", methodHC = "ward.D2")
data_Hc2 <- TCGAanalyze_Clustering(tabDF = datFilt,
                                   method = "consensus",
                                   methodHC = "ward.D2")        # time = 207.389 
# deciding number of tree to cuts
cut.tree <-4
paste0(c("EC"),(1:cut.tree))

## consensusClusters contains barcodes for 4 groups
ans <- hclust(ddist <- dist(datFilt), method = "ward.D2")
hhc <- data_Hc2[[4]]$consensusTree
consensusClusters<-data_Hc2[[4]]$consensusClass
sampleOrder <- consensusClusters[hhc$order]

consensusClusters <- as.factor(data_Hc2[[4]]$clrs[[1]])
names(consensusClusters) <- attr(ddist, "Labels")
names(consensusClusters) <- substr(names(consensusClusters),1,12)

# adding information about gropus from consensus Cluster in clinical data
dataClin <- cbind(dataClin, groupsHC = matrix(0,nrow(dataClin),1))
rownames(dataClin) <- dataClin$bcr_patient_barcode

for( i in 1: nrow(dataClin)){
    currSmp <- dataClin$bcr_patient_barcode[i]
    dataClin[currSmp,"groupsHC"] <- as.character(consensusClusters[currSmp])
}

TCGAanalyze_survival(data = dataClin,
                     clusterCol = "groupsHC",
                     main = "TCGA kaplan meier survival plot from consensus cluster",
                     height = 10, 
                     width=10,
                     legend = "RNA Group",
                     labels=paste0(c("EC"),(1:cut.tree)),
                     color = as.character(levels(consensusClusters)),
                     filename = "figure4surv.png")

```

The figures resulted from the code above are shown below.
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure4cluster.png")
grid.raster(img)
```

```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure4surv.png")
grid.raster(img)
```

```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure4heatmap.png")
grid.raster(img)
```


## Case study n. 3: Integration of methylation and expression for COAD

In recent years, it was discovered that there is a relationship between 
DNA methylation and gene expression. The study of the relationship between the 
two is often difficult to accomplish.

This case study will show the steps to conduct a study of the relationship
between the two types of data.

First we downloaded COAD methylation data for HumanMethylation27k and HumanMethylation450k
platforms, and COAD expresison data for IlluminaGA_RNASeqV2. 

TCGAbiolinks adds by default the classifations already published by researchers.
We will use this classification to do our examples. 
We selected the groups CIMP.L and CIMP.H to do a expression and DNA methylation comparison.

Firstly, we do a DMR (different methylated region) analysis, which will give the 
difference of DNA methylation for the probes of the groups and their significance value.
The ouput can be seen by a volcano plot.

Secondly, we do a DEA (diffential expression analysis) which will give the fold change 
of gene expression and their significance value.

Finally, using both previous analysis we do a starburst plot to select the genes 
that  are Candidate Biologically Significant.

Observation: over the time, the number of samples has increased and the clinical data updated.
We used only the samples that had a classification in the examples.



```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
#-----------------------------------
# STEP 1: Search, download, prepare |
#-----------------------------------
# 1.1 - Methylation
# ----------------------------------
query.met <- TCGAquery(tumor = c("coad"),
                       platform = c("HumanMethylation27",
                                    "HumanMethylation450"),
                       level = 3)

TCGAdownload(query.met, path = "/dados/ibm/comparing/biolinks/coad/")

coad.met <- TCGAprepare(query = query.met,
                        dir = "/dados/ibm/comparing/biolinks/coad/",
                        save = TRUE, filename = "metcoad.rda",reannotate = TRUE)

#-----------------------------------
# 1.2 - Expression
# ----------------------------------

coad.query.exp <- TCGAquery(tumor = "coad",
                       platform = "IlluminaGA_RNASeqV2",
                       level = 3)

TCGAdownload(coad.query.exp,
             path = "/dados/ibm/comparing/biolinks/RNA/",
             type = "rsem.genes.results")

coad.exp <- TCGAprepare(query = coad.query.exp,
                        dir = "/dados/ibm/comparing/biolinks/RNA/",
                        type = "rsem.genes.results",
                        save = T,filename = "coadexp.rda")


# removing the samples without classification
coad.met <- subset(coad.met,select = !(colData(coad.met)$methylation_subtype %in% c(NA)))

#--------------------------------------------
# STEP 2: Analysis
#--------------------------------------------
# 2.1 - Mean methylation of samples
# -------------------------------------------
TCGAvisualize_meanMethylation(coad.met,
                              groupCol = "methylation_subtype",subgroupCol = "hypermutated",
                              group.legend  = "Groups", subgroup.legend = "hypermutated",
                              filename = "coad_mean.pdf")

#------------------------------------------
# 4.2 - Volcano plot
# -----------------------------------------

coad.aux <- subset(coad.met, select = colData(coad.met)$methylation_subtype %in% c("CIMP.L","CIMP.H"))

# na.omit
coad.aux <- subset(coad.aux,subset = (rowSums(is.na(assay(coad.aux))) == 0))

# Volcano plot
coad.aux <- TCGAanalyze_DMR(coad.aux, groupCol = "methylation_subtype",group1 = "CIMP.H",
                       group2="CIMP.L", p.cut = 10^-8, diffmean.cut = 0.25, legend = "State",
                       filename = "coad_CIMPHvsCIMPL_metvolcano.pdf")

save(coad.aux,file = "coad_pvalue.rda")

#------------------------------------------
# 4.4 - Starburst plot
# -----------------------------------------

coad.exp.aux <- subset(coad.exp, select = colData(coad.exp)$methylation_subtype %in% c("CIMP.H","CIMP.L"))

idx <- colData(coad.exp.aux)$methylation_subtype %in% c("CIMP.H")
idx2 <- colData(coad.exp.aux)$methylation_subtype %in% c("CIMP.L")


dataPrep <- TCGAanalyze_Preprocessing(object = coad.exp.aux,cor.cut = 0.6)
dataNorm <- TCGAanalyze_Normalization(tabDF = dataPrep,
                                      geneInfo = geneInfo,
                                      method = "gcContent")

dataFilt <- TCGAanalyze_Filtering(tabDF = dataNorm, qnt.cut = 0.25)

dataDEGs <- TCGAanalyze_DEA(mat1 = dataFilt[,idx],
                            mat2 = dataFilt[,idx2],
                            Cond1type = "Normal",
                            Cond2type = "Tumor",
                            method = "glmLRT")

TCGAVisualize_volcano(dataDEGs$logFC,dataDEGs$FDR,
                      filename="Case3_volcanoexp.pdf",
                     label= c("1" = "Not Significant",  "2" = "Up regulared",
                        "3" = "Down regulated"),
                     x.cut=4, names= rownames(dataDEGs),
                     xlab =" Log2 fold change",
                     legend="State",title = "Volcano plot (CIMP.l vs CIMP.H)")

starburst <- TCGAvisualize_starburst(coad.aux, dataDEGs,"CIMP.H","CIMP.L",
                                     met.p.cut = 10^-5, exp.p.cut=10^-5,
                                     names = TRUE)
```

The figures resulted from the code above are shown below.
```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure5exp.png")
grid.raster(img)
```

```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure5met.png")
grid.raster(img)
```

```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("figure5star.png")
grid.raster(img)
```

## Case study n. 4: Elmer pipeline - KIRC

One of the biggest problems related to the study data is the preparation phase,
which often consists of successive 
steps in order to prepare it to a format acceptable by certain algorithms and software.

With the object of assisting users in this arduous step, TCGAbiolinks offers in 
data preparation stage, the toPackage argument, which aims to prepare the data 
in order to obtain the correct format for different packages.

An example of another package to perform analysis is ELMER. We will present this
case study the study KIRC by TCGAbiolinks and ELMER integration.

```{r,eval=FALSE,echo=TRUE,message=FALSE,warning=FALSE}
#-----------------------------------
# STEP 1: Search, download, prepare |
#-----------------------------------
# Step 1.1 download methylation data|
# -----------------------------------
path <-  "."
query <- TCGAquery(tumor = "KIRC",level = 3, platform = "HumanMethylation450")
TCGAdownload(query, path = path)
kirc.met <- TCGAprepare(query,dir = path, save = TRUE,
                        filename = "metKirc.rda",
                        toPackage = "ELMER")

# Step 1.2 download expression data
query.rna <- TCGAquery(tumor="KIRC",level=3, platform="IlluminaHiSeq_RNASeqV2")
TCGAdownload(query.rna,path=path,type = "rsem.genes.normalized_results")

kirc.exp <- TCGAprepare(query.rna, dir=path, save = TRUE,
                        type = "rsem.genes.normalized_results",
                        filename = "expKirc.rda",toPackage = "ELMER")

#-----------------------------------
# STEP 2: ELMER integration         |
#-----------------------------------
# Step 2.1 prepare mee object       |
# -----------------------------------
library(ELMER)
library(parallel)

geneAnnot <- txs()
geneAnnot$GENEID <- paste0("ID",geneAnnot$GENEID)
geneInfo <- promoters(geneAnnot,upstream = 0, downstream = 0)
probe <- get.feature.probe()
mee <- fetch.mee(meth = kirc.met, exp = kirc.exp, TCGA = TRUE,
                 probeInfo = probe, geneInfo = geneInfo)

#--------------------------------------
# STEP 3: Analysis                     |
#--------------------------------------
# Step 3.1: Get diff methylated probes |
#--------------------------------------
Sig.probes <- get.diff.meth(mee ,cores=detectCores(),
                            dir.out ="kirc",diff.dir="hypo",
                            pvalue = 0.01)


#--------------------------------------------------
# Step 3.2: Identifying putative probe-gene pairs |
#--------------------------------------------------
# Collect nearby 20 genes for Sig.probes
nearGenes <- GetNearGenes(TRange=getProbeInfo(mee, probe=Sig.probes),
                          cores=detectCores(),
                          geneAnnot=getGeneInfo(mee))

# Identify significant probe-gene pairs
Hypo.pair <- get.pair(mee=mee,
                      probes=Sig.probes,
                      nearGenes=nearGenes,
                      permu.dir="./kirc/permu",
                      dir.out="./kirc/",
                      cores=detectCores(),
                      label= "hypo",
                      permu.size=10000,
                      Pe = 0.01)

Sig.probes.paired <- fetch.pair(pair=Hypo.pair,
                                probeInfo = getProbeInfo(mee),
                                geneInfo = getGeneInfo(mee))

#-------------------------------------------------------------
# Step 3.3: Motif enrichment analysis on the selected probes |
#-------------------------------------------------------------
enriched.motif <- get.enriched.motif(probes=Sig.probes.paired,
                                     dir.out="kirc", label="hypo",
                                     background.probes = probe$name)

#-------------------------------------------------------------
# Step 3.4: Identifying regulatory TFs                        |
#-------------------------------------------------------------
TF <- get.TFs(mee=mee,
              enriched.motif=enriched.motif,
              dir.out="kirc",
              cores=detectCores(), label= "hypo")

```

From this analysis it is possible to verify the relation between a probe 
and nearby genes. The result of this is show by the ELMER scatter plot.

```{r, fig.width=6, fig.height=4, echo = FALSE, fig.align="center",hide=TRUE, message=FALSE,warning=FALSE}
library(png)
library(grid)
img <- readPNG("case4_elmer.png")
grid.raster(img)
```
******
### Session Information
******
```{r sessionInfo}
sessionInfo()
```

# References
