---
title: "Introduction"
date: "`r BiocStyle::doc_date()`"
vignette: >
  %\VignetteIndexEntry{"Introduction"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(progress = FALSE)
```

TCGAbiolinks is able to access The National Cancer Institute (NCI) Genomic Data Commons (GDC) thorough its  
[GDC Application Programming Interface (API)](https://gdc.cancer.gov/developers/gdc-application-programming-interface-api) 
to search, download and prepare relevant data for analysis in R. 


# Installation

## R
You can install the stable version from [Bioconductor](http://bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html), or the development version using:

* Stable version:
```{r message=FALSE, warning=FALSE, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("TCGAbiolinks")
```

* Development version:
```{r message=FALSE, warning=FALSE, eval=FALSE}
devtools::install_github('BioinformaticsFMRP/TCGAbiolinks')
```

## Docker image

If you prefer to use docker, TCGAbiolinks and TCGAbiolinksGUI are available as Docker image (self-contained environments that contain everything needed to run the software), 
which can be easily run on Mac OS, Windows and Linux systems. 

This [PDF](https://drive.google.com/open?id=0B0-8N2fjttG-QXp5LVlPQnVQejg) show how to install and execute the image.

The image can be obtained from Docker Hub: https://hub.docker.com/r/tiagochst/tcgabiolinksgui/

For more information please check: https://docs.docker.com/ and https://www.bioconductor.org/help/docker/


# Question and issues

Please use [Github issues](https://github.com/BioinformaticsFMRP/TCGAbiolinks/issues) if you want to file bug reports or feature requests.

# Required libraries

The examples in this tutorial use the following libraries: 
```{r message=FALSE, warning=FALSE, include=TRUE}
library(TCGAbiolinks)
library(dplyr)
library(DT)
```

# Session info
```{r}
version
packageVersion("TCGAbiolinks")
```
